---
title: "STAT111 R Bootcamp"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Roodmap

* What is R?
* Basics and fundmaentals
* Data exploration exercises and OHs

## What is R?

* Open-source language for statistical computing and graphics
* Ranked 12th TIOBE index.
* Built for things statisticians care about:
    - Vectors and matrices
    - Linear/non-linear modeling
    - Statistical tests
    - Visualization and more!
* Highly extensible because of packages (and you can write your own, too!)

## R and Stat 111
* Do not worry if you have no coding experience!
* Goal of this workshop is to build comfort looking things up and solving your own problems in R
* Homeworks may involve computation on real/synthetic data to demonstrate inference in practice
    - Other languages are also allowed (e.g. Python)
* At OH, Course staff will primarily provide conceptual support with computational problems, NOT debugging code

## RStudio
* An integrated development environment (IDE) to organize your code, data, plots, files, and more.
* You can even type your homeworks using RMarkdown!


## R Resources
* DataCamp, whose Chief Data Scientist was a Harvard stat concentrator
* R for Data Science , the free online book by Grolemund and Wickham.
* Google and Stackoverflow

## Language Fundamentals
* Functions and documentation
* Variables and data types
* Vectors, matrices, data frames
* Loops, conditionals, vectorization
* Plotting

Not covered but worth reading: Factors, lists, pipes, apply() function

## Functions and Documentation
* Functions take inputs ("arguments") and return an output

\scriptsize
```{r, echo = TRUE}
sum(110, 111, 211)
```
\normalsize
* Sometimes, arguments are optional or carry a default value.
* \textbf{Exercise}: Look up \texttt{rnorm()} and generate 10 i.i.d. N(5, 1) data points. What do \texttt{pnorm(), dnorm()} and \texttt{qnorm()} do?
(Hint: Use ?rnorm to read documentation)

\pause
\scriptsize
```{r, echo = TRUE}
?rnorm
rnorm(n = 10, mean = 5, sd = 1)
pnorm(0.5)
dnorm(0.5)
qnorm(0.5)
```

## Writing Your Own Functions
\scriptsize
```{r, echo = TRUE}
SayHello <- function(name = "world"){
  # This is a comment
  # Use ?cat to see what cat do
  # function will "return" or output the value of the last line
  # also can use "return()"
  cat("hello", name)
}
SayHello()
SayHello("Joe and Susan")
```

## Variables
* Often store values in "buckets" called variables. 
    - Represent values without having to retype them
    - Update a variable if the value changes
    
\scriptsize
```{r, echo = TRUE}
# Assign or update using "<-"
# "=" also works
mu <- 2.56
y <- rnorm(1, mean = mu)
pnorm(y, mean = mu)
# update
mu <- mu + 1
pnorm(y, mean = mu)
print(y)
```


## Basic Data Types
* Values in R are generally one of the following types: 
    - Logical:  \texttt{TRUE, FALSE}
    - Integer:  \texttt{110, 111}
    - Decimal:  \texttt{3.14}
    - Complex: \texttt{3+2i}
    - Character:  \texttt{"joe", "susan"}
* Use \texttt{class()} to check a value’s type.
\pause
* Careful -- not all data types play well together!

\scriptsize
```{r, echo = TRUE}
# This won’t work
# "hello" + 1
# This is okay
1 + 3+2i + 2.72
# This is okay, too; logicals are coerced into 1’s and 0’s
4 + TRUE + FALSE
```

## Useful Operators for Numerics
* Operators
    - Basic operations: +, -, *, /
    - Exponents: ^
    - Modulo: %%
    - Comparisons: <, <=, >, >=, ==, !=
    - Other functions: exp(), log(), sin(), cos()
* \textbf{Exercise}: Predict results for the following computations: 

\pause
\scriptsize
```{r, echo = TRUE, eval=F}
110 < 111
exp(log(-1))
4 + 12 / 6 %% 3
```

## Useful Operators for Numerics
* Operators
    - Basic operations: +, -, *, /
    - Exponents: ^
    - Modulo: %%; Quotient: %/%
    - Comparisons: <, <=, >, >=, ==, !=
    - Other functions: exp(), log(), sin(), cos()
* \textbf{Exercise}: Predict results for the following computations: 

\scriptsize
```{r, echo = TRUE, eval=T}
110 < 111
exp(log(-1))
4 + 12 / 6 %% 3
```

## Vectors
* Like in math, a 1d collection of elements of the same type.

\scriptsize
```{r, echo = TRUE, eval=T}
# Create a vector
my_vector <- vector("numeric", length = 3)
my_vector
# my_vector <- numeric(length = 3)
# Combine vectors
c(my_vector, -7, 32)
# Vector arithmetic
my_vector + 1
my_vector + c(4, 2, 5)
```

## Vectors: Naming and Accessing Elements

\scriptsize
```{r, echo = TRUE, eval=T}
enrollment <- c(586, 328, 51)
# Give elements names
names(enrollment) <- c("Stat 110", "Stat 111", "Stat 211")
# Access Stat 111 enrollment in a few different ways
enrollment[2]; enrollment["Stat 111"];
enrollment[c(FALSE, TRUE, FALSE)]
```
\textbf{Exercise}: How would you get the enrollment of Stat 110 AND Stat 111?
\pause
\tiny
```{r, echo = TRUE, eval=T}
enrollment[c(1, 2)]
enrollment[c("Stat 110", "Stat 111")]
```

## Vectors: Useful Functions
\scriptsize
```{r, echo = TRUE, eval=T}
# The colon operator or seq() function yield sequences.
my_vector <- 1:10
my_other_vector <- seq(1, 10, by = 2)

# Other useful functions
mean(my_vector)
var(my_vector)
summary(my_vector)
table(my_vector)
```

## Matrices
* Like in math, a 2d collection of elements of the same type.

\scriptsize
```{r, echo = TRUE, eval=T}
# Create a matrix
my_matrix <- matrix(1:6, nrow = 2, ncol = 3, byrow = TRUE)
# Matrix arithmetic
my_matrix * 2
# Matrix multiplication and transpose
my_matrix %*% t(my_matrix)
```

## Matrices
\textbf{Exercise}: Try \texttt{my\_matrix * my\_matrix} ?
\pause
\scriptsize
```{r, echo = TRUE, eval=T}
# Create a matrix
my_matrix * my_matrix
```

## Matrices: Naming and Accessing Elements

\scriptsize
```{r, echo = TRUE, eval=T}
enrollment = matrix(c(586, 620, 328, 355, 51, 44), nrow = 2)
rownames(enrollment) = c("2018", "2019")
colnames(enrollment) = c("Stat 110", "Stat 111", "Stat 211")
enrollment
```
\normalsize
* Predict the following

\scriptsize
```{r, echo = TRUE, eval=F}
enrollment[1, 2]
enrollment["2020", "Stat 110"]
enrollment[2, ]
```


## Matrices: Useful Functions

\scriptsize
```{r, echo = TRUE, eval=F}
# Merge matrices together; also see rbind()
enrollment = cbind(enrollment, "Stat 139" = c(152, 175))
```
```{r, echo = TRUE, eval=F}
# You can guess what these do
dim(); nrow(); ncol()
rowSums(); colSums()
```

## Matrices: Useful Functions

\scriptsize
```{r, echo = TRUE, eval=T}
# Merge matrices together; also see rbind()
enrollment = cbind(enrollment, "Stat 139" = c(152, 175))
```
```{r, echo = TRUE, eval=F}
# You can guess what these do
dim(enrollment); nrow(enrollment); ncol(enrollment)
rowSums(enrollment); colSums(enrollment)
```

## Data Frames
* Like matrices, but columns can be of different types.

\scriptsize
```{r, echo = TRUE, eval=T}
# Create a data frame
my_df <- data.frame("Dept" = c("Stat", "CS", "Gov"),
 "Enrollment" = c(177, 244, 221))
```
```{r, echo = TRUE, eval=F}
# More commonly, open a file
my_df <- read.table("filename")
my_df <- read.csv("filename")
```
```{r, echo = TRUE, eval=T}
# Preview data
head(my_df); tail(my_df); 
# View(my_df); str(mu_df)
```

## Data Frames
\scriptsize
```{r, echo = TRUE, eval=T}
my_df 
```
* Predict the following:
```{r, echo = TRUE, eval=F}
subset(my_df, Enrollment > 200)
my_df[nrow(my_df):1, ]
my_df[sample(nrow(my_df)), ]
```
\pause
```{r, echo = F, eval=T}
subset(my_df, Enrollment > 200)
my_df[nrow(my_df):1, ]
```

## Conditionals: If … Then … Else ...
\scriptsize
```{r, echo = T, eval=T}
x <- rnorm(1)
x
if (x > 0) {
	print("You got a positive number!")
} else {
	print("You got a negative number!")
}
```

## Loops: Repeat Chunks of Code
* For loops vs while loops

\scriptsize
```{r, echo = T, eval=T}
total <- 0
for (i in 1:10) {
	total <- total + i
}
print(total)

total <- 0
i <- 1
while (i <= 10) {
	total <- total + i
	i <- i + 1
}
print(total)
```


## Loops vs. Vectorization
* Vectoriztion: 
      - Avoid loops/conditionals
      - Vectorized functions that apply an operation to an entire vector and return a vector.
      - Cleaner and computationally faster!
      - See the apply() function for more!

\scriptsize
```{r, echo = T, eval=T}
sum(1:10)
```
```{r, echo = T, eval=T}
enrollment
apply(enrollment, 2, mean)
```

## Visualization: Plotting with ggplot2
* R has built-in plotting called base graphics, which follows a “pen-and-paper” model
* ggplot2 is more modular, layering elements individually
* Google: "ggglot2 cheatsheet""
* Highly recommended: Chapter 3 of R for Data Science
* Install ggplot2 package
```{r, echo = T, eval=F}
install.packages("ggplot2")
library(ggplot2)
```
```{r, echo = F, eval=T}
library(ggplot2)
```

## ggplot2: The Grammar of Graphics
\scriptsize
```{r, echo = T, eval=F}
graphic <- ggplot(data = <DATA>) + 
  	<GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```
```{r, echo = T, eval=F}
# Try explore "mgp" data first
?mgp
str(mpg)
```
```{r, echo = T, eval=T, fig.width=10, fig.height=4}
# scatterplot: engine displacement vs. highway miles per gallon
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

## ggplot2: The Grammar of Graphics
* Layer geoms together
* Mappings can be specified globally or locally

\scriptsize
```{r, echo = T, eval=T, fig.width=10, fig.height=4}
# Then add a color aesthetic for class.
# Other aesthetics you can try include shape, size, and alpha.
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
```

## ggplot2: The Grammar of Graphics
\scriptsize
```{r, echo = T, eval=T, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
# histogram
ggplot(data = mpg, mapping = aes(x = displ, fill = class)) +
geom_histogram()
# look up geom_histogram, adjust binwidth, boundary...
```

## Exercises and Data Exploration
* Logs of all Bluebikes trips in December 2018
* https://canvas.harvard.edu/files/7228708/
* Attempt the exercises on the handout and ask a TF if you need help!
